---
title: "Housing Prices"
author: "Ozan Aygun"
date: "11/26/2017"
output: 
   html_document:
        toc: true
        number_sections: true
        depth: 4
        theme: cerulean
        highlight: tango
        df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "markup", fig.align = "center",
                      fig.width = 5, fig.height = 4,message=FALSE,warning=FALSE)
```


# Introduction and summary

# Loading the data

```{r, echo=FALSE}
setwd("~/Desktop/2016/Data_science/Kaggle/House_prices_Kaggle_competition")
testURL <- "https://storage.googleapis.com/kaggle-competitions-data/kaggle/5407/test.csv?GoogleAccessId=competitions-data@kaggle-161607.iam.gserviceaccount.com&Expires=1512002091&Signature=Iyzcu0kfgpwTx2oO0GMUyfGbPp5WM7S13a%2Bbt7SdJNTXo%2FOtpQgoW9k%2BrV65bjG6vhNIohOVXYw10JzvoDeHw4w07Ymuwyv0Qt5aU8KAxV4ZoJ2Gc3uvems%2Bw6WCe4xCFp5o8m20NtTlLdccMNWXe%2FGGFFcTNYFpFvx9G5dmeTwjlaBOsyDdm5onF4G2PG4Qi5GJfuZ213808IofjxArwMEyt4aWiP5%2B0f26DiZxUoNniPmUESrkGg47b78dPL9ip6%2FpnTZJsgPnk8KI8KQyq%2BgzZRTJdbwCGiopDiGz9VQERCKHig4W3eniPYtdByXbtT8myVRMdU%2BW7hw0WgTwkg%3D%3D"
download.file(testURL, destfile = "./test.csv")

trainURL <-"https://storage.googleapis.com/kaggle-competitions-data/kaggle/5407/train.csv?GoogleAccessId=competitions-data@kaggle-161607.iam.gserviceaccount.com&Expires=1512002321&Signature=BXfAxWBLVg%2BCnlx70kkKU7FBqLrB0mBXkvSCxMN10yORFGjHxWnTvXl%2FDCc%2Bcx10URdODg%2FW0s6RdGLNxrOyImlAMTSRNBk1jZJWCsv%2BY0TfRIKjb0R9ebD44m1b07vTDIdIE%2BnF2x4zBLrwR%2FMdCLOA9VQaf0uGYaPKD%2BewwzsU5i3sTr0VK03SBn6NUlst9W8g69irdAoNqjuDHhCHTwyQ4dQ6JxNVMeq9ZOL33XlXxuhfVaUvL1bwSUc8OcW%2BBCKlFPfzfjR8AKDx4D5fAS2Yn6cmivXHG0xJswSkdqRvHiDlHpUaAIEh4RzmF5rozEOZ1Btf6TIs5Ka9atirYg%3D%3D"
download.file(trainURL, destfile = "./train.csv")
```

```{r}
train <- read.csv("train.csv"); test <- read.csv("test.csv")
```

## Summarizing the data
```{r}
summary(train)
```

The outcome variable SalePrice is a complete variable, but there are a number of predictors with many missing values. Median house price is $163,000.

```{r}
par(mfrow = c(1,2))
boxplot(train$SalePrice,main = "SalePrice")
boxplot(log10(train$SalePrice), main = "Log10(SalePrice)")
```

## Understanding the impact of missing data

Which features have more than 30% missing data?

```{r}
na.status <- is.na(train)
na.sum <- apply(na.status,2,sum)
names(na.sum) <- colnames(train)
mostly_missing <- which(na.sum > (0.3 * nrow(train)))
na.sum[mostly_missing]
```

5 variables have substantial amount of missing data. Do they have impact on the house prices?

###Alley

Based on the data description: Type of alley access to property

       Grvl     Gravel
       Pave     Paved
       NA       No alley access
```{r}
library(ggplot2)

ggplot(data = train, aes(x = Alley, y = log(SalePrice), fill= Alley))+
        geom_boxplot()
```
Alley seems to be important. Gravel homes have significantly lower price. Convert NAs to "NoAlley" to make more sense.

```{r}
train$Alley <- as.character(train$Alley)
train$Alley[which(is.na(train$Alley))] <- "NoAlley"
train$Alley <- factor(train$Alley)

#Also transform the test set
test$Alley <- as.character(test$Alley)
test$Alley[which(is.na(test$Alley))] <- "NoAlley"
test$Alley <- factor(test$Alley)
```

###FireplaceQu

Based on the data desxription:

FireplaceQu: Fireplace quality

       Ex       Excellent - Exceptional Masonry Fireplace
       Gd       Good - Masonry Fireplace in main level
       TA       Average - Prefabricated Fireplace in main living area or Masonry Fireplace in basement
       Fa       Fair - Prefabricated Fireplace in basement
       Po       Poor - Ben Franklin Stove
       NA       No Fireplace

```{r}

ggplot(data = train, aes(x = FireplaceQu, y = log10(SalePrice), fill= FireplaceQu))+
        geom_boxplot()

```

Let's look at through a simple linear model:

```{r}
par(mfrow=c(1,2))
hist(train$SalePrice)
hist(log10(train$SalePrice)) # Looks more gaussian



FirePlaceFit <- lm(log10(train$SalePrice) ~ FireplaceQu, data = train)
summary(FirePlaceFit)
par(mfrow=c(2,2))
plot(FirePlaceFit)
```
It appears that all levels of this feature also have significantly different mean Sales prices. Again, converting NAs to NoFireplace to be more consistent.

```{r}
train$FireplaceQu <- as.character(train$FireplaceQu)
train$FireplaceQu[which(is.na(train$FireplaceQu))] <- "NoFireplace"
train$FireplaceQu <- factor(train$FireplaceQu)

#Also transform the test set
test$FireplaceQu <- as.character(test$FireplaceQu)
test$FireplaceQu[which(is.na(test$FireplaceQu))] <- "NoFireplace"
test$FireplaceQu <- factor(test$FireplaceQu)
```



###PoolQC






